\large{\bf Quantitative genetics with stable laws }

### 2023-06-21 writeup.tex
minor edits

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\mL}{\mathcal{L}}
\newcommand{\ip}[2]{\left\langle #1, #2 \right\rangle}
\newcommand{\ie}{\textit{i.e.,}\,}
\newcommand{\eg}{\textit{e.g.,}\,}
\newcommand{\nb}{\textit{n.b.,}\, }
\newcommand{\defn}{:=}

@@

@@

Let $X$ be a population of size $N$, recorded as the empirical distribution on phenotypes
\begin{align*}
    X = \frac{1}{N} \sum_{i=1}^N \delta_{x_i} .
\end{align*}
This evolves as a Moran process under the infinitesimal model
with death rates depending on state $x$.
Concretely, an individual whose phenotype is $x$
dies at rate $1 + \mu(x)$
and is replaced by a new individual of phenotype
\begin{align*}
  \frac{y + z}{2} + \xi ,
\end{align*}
where $y$ and $z$ are independent draws from $X$
and $\xi$ is an independent $N(0, \eta^2)$.
Let $\nu(X)$ denote the distribution of $\frac{y + z}{2} + \xi$.
This has generator
\begin{align*}
    \mL \Phi(X)
    &=
    \int \int \int \int
    \left( \Phi(X + \frac{1}{N} \delta_{(y+z)/2 + \xi} - \frac{1}{N} \delta_x) - \Phi(X) \right)
    (1 + \mu(x)) X(dx) X(dy) X(dz) e^{-(\xi/\eta)^2/2} d\xi \\
    &=
    \int \int
    \left( \Phi(X + \frac{1}{N} \delta_u - \frac{1}{N} \delta_x) - \Phi(X) \right)
    (1 + \mu(x)) X(dx) \nu(X)(du) .
\end{align*}

For a test function $\phi$, setting $\Phi(X) = \ip{\phi}{X}$
we get the first moment
\begin{align*}
    \mL \ip{\phi}{X}
    &=
    \frac{1}{N} \int \int \int \int
    \left( \phi \left( \frac{y+z}{2} + \xi \right)
         - \phi(x) \right)
    (1 + \mu(x)) X(dx) X(dy) X(dz) e^{-(\xi/\eta)^2/2} d\xi  \\
    &=
    \frac{1}{N} \ip{\phi}{\nu(X)}\ip{1+\mu}{X}
    - \frac{1}{N} \ip{\phi}{(1+\mu) X} .
\end{align*}
Note that $\ip{\phi}{\partial_{\nu(X)} X} = \ip{\phi}{\nu(X)}$,
so that this is a first derivative in the direction of
$\ip{1+\mu}{X} \nu(X) - (1+\mu) X$.

@@

\paragraph{Stable measures}
If $Z$ is a fixed point of this flow,
then $\nu(Z)\ip{1+\mu}{Z} = (1+\mu)Z$.
Any measure $Z$ that is a fixed point has the property that
(the average of two independent draws plus segregation noise)
is equal in distribution to
(a $1+\mu$-weighted draw).
Concretely, with $A$, $B$, and $C$ independent draws from $Z$,
$$\begin{aligned}
    \E\left[ f\left( \frac{A+B}{2} + \xi \right) \right]
    &=
    \frac{\E\left[ f(C) (1+\mu(C)) \right]}{\E[1+\mu(C)]} .
\end{aligned}$$

@@

We now analyse the component arising from the noise $\xi$.  First, we consider the case when $\Psi(\theta)$ is twice continuously differentiable, so $\xi$ has finite mean $m = \Psi'(0)$ and variance $\eta^{2 = \Psi''(0)}$.   Taylor's theorem then tells us that for $\theta$ sufficiently small, 
\[
	\Psi(\theta) = m\theta + \frac{\eta^{2}}{2} \theta^{2} + r(\theta)\theta^{2},
\]
where $r(\theta) \to 0$ as $\theta \to 0$.  Then,
\begin{align*}
	 \sum_{k=1}^{n-1}  {\textstyle 2^{k} \Psi\left(\frac{\theta}{2^{k}}\right)}
	 &=  \sum_{k=1}^{n-1} 2^{k} {\textstyle \left(m \frac{\theta}{2^{k}} - \frac{\eta^{2}}{2}  \frac{\theta^{2}}{2^{2k}}
		+ r\left(\frac{\theta}{2^{k}}\right)\frac{\theta^{2}}{2^{2k}}\right)}\\
	&= (n-1) m - \eta^{2} \theta^{2} (1 - 2^{-n+1}) 
		+ \sum_{k=1}^{n-1} {\textstyle r\left(\frac{\theta}{2^{k}}\right)\frac{\theta^{2}}{2^{k}}}
 \end{align*}

@@

Now, consider the case when $1+ \mu(x) = e^{a x^{2}}$.  In this case there is a Gaussian solution: suppose $Z$ is Gaussian with mean $m$ and variance $\sigma^{2}$.  Then,
\[
	\phi(\theta) = e^{i m \theta - \frac{1}{2} \sigma^{2} \theta^{2}}.
\]
Moreover, provided $a < \frac{1}{2\sigma^{2}}$, we can explicitly compute $\frac{\mathbb{E}\left[e^{i \theta Z}(1+\mu(Z))\right]}{\mathbb{E}[1+\mu(Z)]}$  to conclude that
if $\xi$ is Gaussian with mean zero and variance $\eta^2$,
then equation~\eqref{CHARACTERISTIC} is
\comment{something not right here? need a factor of 2 on the left?}
\[
	e^{i m \theta - \frac{1}{2} \left(\frac{\sigma^{2}}{2} + \eta^{2}\right)\theta^{2}}
	= e^{i \frac{m}{1-2 a \sigma^{2}}\theta 
		- \frac{1}{2} \frac{\sigma^{2}}{1-2 a \sigma^{2}}\theta^{2}},
\]
which, equating real and complex components, has a solution provided $m = 0$ and
\[
	\sigma^{2} = \frac{\sqrt{16 a^{2} \eta^{4} + 24 a \eta^{2}+1} - 4a \eta^{2} -1}{4a} 
	=  2\eta^{2}(1-4a\eta^{2}) + O(a^{2}).
\]

@@

In Section~\ref{sec:stabilizing_selection}
we have found two explicit conditions in which
if the form of segregation noise and stabilizing selection
satisfy certain conditions,
then the limiting phenotype distribution
will be either Gaussian (Subsection~\ref{sec:stabilizing_gaussian}
or Cauchy (Subsection~\ref{sec:stabilizing_cauchy}.
Furthermore, we expect based on
\citet{barton2017infinitesimal} and arguments above
that the distribution of segreation noise
is determined by the tail exponent of the distribution of effect sizes
(under appropriate assumptions; e.g., that most mutations are of small effect).
However, it is not clear how selection would affect these arguments,
especially in the stable case where some number of mutations are expected to
\emph{not} have small effect (and therefore such mutations will not be nearly neutral).

Therefore,
to study how this works in practice, we turn to simulation.
We used SLiM v3 \comment{(cite)}
to simulate a discrete-time approximation to the continuous-time Moran model described above
as follows.
Individuals are diploid and sexual;
the genome is of lenth $10^8$bp with a uniform recombination rate of $10^{-8}$ crossovers per bp;
mutations occur at a rate of $m$ per bp.
Each new mutation is independently assigned an ``effect''
drawn from the ``effect size distribution'' (and will be either Gaussian or Cauchy).
The phenotype of each individual is determined additively
by summing the effects of all mutations they carry across both chromosomes.
Let $f(z) = 1 + \mu(z)$
be the ``fitness'' of an individual with phenotype $z$,
and let $dt$ be a small constant (we took $dt=0.01$).
Then, at each time step
each individual dies with probability $1 - \exp(-dt f(z))$;
if there are $k$ such individuals chosen to die in a given time step,
then we choose $k$ individuals uniformly and without replacement
to produce new offspring;
for each such reproduction event the parent chooses a mate uniformly at random from the population.
This maintains the population at a fixed size, $N$.
(Note that individuals may self and individuals chosen to die may also reproduce,
but in a large population these details should be unimportant.)

>>>>

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\mL}{\mathcal{L}}
\newcommand{\ip}[2]{\left\langle #1, #2 \right\rangle}
\newcommand{\ie}{\textit{i.e.,}\,}
\newcommand{\eg}{\textit{e.g.,}\,}
\newcommand{\nb}{\textit{n.b.,}\, }
\newcommand{\defn}{:=}

@@

\comment{Proposal: change $1 + \mu$ to $f$ (or, something else, but a single letter).}

@@

Let $X$ be a population of size $N$, recorded as the empirical distribution on phenotypes
\begin{align*}
    X = \frac{1}{N} \sum_{i=1}^N \delta_{x_i} .
\end{align*}
So, $X$ is a point measure of total mass 1 on $\R$.
This evolves as a Moran process under the infinitesimal model
with death rates depending on state $x$.
Concretely, an individual whose phenotype is $x$
dies at rate $1 + \mu(x)$
and is replaced by a new individual of phenotype
\begin{align*}
  \frac{y + z}{2} + \xi ,
\end{align*}
where $y$ and $z$ are independent draws from $X$
and $\xi$ is an independent draw from the segregation noise distribution.
Let $p_\xi$ denote the distribution of $\xi$,
and $\nu(X)$ denote the distribution of $\frac{y + z}{2} + \xi$.
This has generator
\begin{align*}
    \mL \Phi(X)
    &=
    \int \int \int \int
    \left( \Phi\left( X + \frac{1}{N} \delta_{(y+z)/2 + \xi} - \frac{1}{N} \delta_x \right) - \Phi(X) \right)
    (1 + \mu(x)) X(dx) X(dy) X(dz) p_\xi(d\xi) \\
    &=
    \int \int
    \left( \Phi\left(X + \frac{1}{N} \delta_u - \frac{1}{N} \delta_x\right) - \Phi(X) \right)
    (1 + \mu(x)) X(dx) \nu(X)(du) .
\end{align*}

For a test function $\phi$, setting $\Phi(X) = \ip{\phi}{X} := \int \phi(x) X(dx)$
we get the first moment
\begin{align*}
    \mL \ip{\phi}{X}
    &=
    \frac{1}{N} \int \int \int \int
    \left( \phi \left( \frac{y+z}{2} + \xi \right)
         - \phi(x) \right)
    (1 + \mu(x)) X(dx) X(dy) X(dz) p_\xi(d\xi) \\
    &=
    \frac{1}{N} \ip{\phi}{\nu(X)}\ip{1+\mu}{X}
    - \frac{1}{N} \ip{\phi}{(1+\mu) X} .
\end{align*}
Note that for $Y$ another finite measure on $\R$,
the Fr\'echet derivative \comment{(is that the right name?)} of $X$ in the direction of $Y$
satisfies $\ip{\phi}{\partial_Y X} = \lim_{\epsilon \to 0} \epsilon^{-1} (\ip{\phi}{X + \epsilon Y} - \ip{\phi}{X}) = \ip{\phi}{Y}$,
so that this is a first derivative in the direction of the measure
$$ \ip{1+\mu}{X} \nu(X) - (1+\mu) X . $$

@@

\paragraph{Stable measures}
If $Z$ is a fixed point of this flow,
then $\nu(Z)\ip{1+\mu}{Z} = (1+\mu)Z$.
In other words, any measure $Z$ that is a fixed point has the property that
(the average of two independent draws plus segregation noise)
is equal in distribution to
(a $1+\mu$-weighted draw).
Concretely, with $A$, $B$, and $C$ independent draws from $Z$,
\begin{align} \label{eqn:ABC_condition}
    \E\left[ f\left( \frac{A+B}{2} + \xi \right) \right]
    &=
    \frac{\E\left[ f(C) (1+\mu(C)) \right]}{\E[1+\mu(C)]} .
\end{align}

@@

We now analyse the component arising from the noise $\xi$.  First, we consider the case when $\Psi(\theta)$ is twice continuously differentiable, so $\xi$ has finite mean $m = \Psi'(0)$
and variance $\eta^{2} = \Psi''(0)$.   Taylor's theorem then tells us that for $\theta$ sufficiently small, 
\[
	\Psi(\theta) = m\theta + \frac{\eta^{2}}{2} \theta^{2} + r(\theta)\theta^{2},
\]
where $r(\theta) \to 0$ as $\theta \to 0$.  Then,
\begin{align*}
	 \sum_{k=1}^{n-1}  {\textstyle 2^{k} \Psi\left(\frac{\theta}{2^{k}}\right)}
	 &=  \sum_{k=1}^{n-1} 2^{k} {\textstyle \left(m \frac{\theta}{2^{k}} - \frac{\eta^{2}}{2}  \frac{\theta^{2}}{2^{2k}}
		+ r\left(\frac{\theta}{2^{k}}\right)\frac{\theta^{2}}{2^{2k}}\right)}\\
	&= (n-1) m - \eta^{2} \theta^{2} (1 - 2^{-n+1}) 
		+ \sum_{k=1}^{n-1} {\textstyle r\left(\frac{\theta}{2^{k}}\right)\frac{\theta^{2}}{2^{k}}}
 \end{align*}

@@

Now, consider the case when $1+ \mu(x) = e^{a x^{2}}$. 
In this case there is a Gaussian solution: suppose $Z$ is Gaussian with mean $m$ and variance $\sigma^{2}$.  Then,
\[
	\phi(\theta) = e^{i m \theta - \frac{1}{2} \sigma^{2} \theta^{2}}.
\]
Moreover, provided $a < \frac{1}{2\sigma^{2}}$, we can explicitly compute $\frac{\mathbb{E}\left[e^{i \theta Z}(1+\mu(Z))\right]}{\mathbb{E}[1+\mu(Z)]}$  to conclude that
if $\xi$ is Gaussian with mean zero and variance $\eta^2$,
then equation~\eqref{eqn:ABC_condition} with $f(x)=e^{i\theta x}$ is
\[
	e^{i m \theta - \frac{1}{2} \left(\frac{\sigma^{2}}{2} + \eta^{2}\right)\theta^{2}}
	= e^{i \frac{m}{1-2 a \sigma^{2}}\theta 
		- \frac{1}{2} \frac{\sigma^{2}}{1-2 a \sigma^{2}}\theta^{2}},
\]
which, equating real and complex components, has a solution provided $m = 0$ and
\[
	\sigma^{2} = \frac{\sqrt{16 a^{2} \eta^{4} + 24 a \eta^{2}+1} - 4a \eta^{2} -1}{4a} 
	=  2\eta^{2}(1-4a\eta^{2}) + O(a^{2}).
\]

@@

In Section~\ref{sec:stabilizing_selection}
we have found two explicit conditions in which
if the form of segregation noise and stabilizing selection
satisfy certain conditions,
then the limiting phenotype distribution
will be either Gaussian (Subsection~\ref{sec:stabilizing_gaussian})
or Cauchy (Subsection~\ref{sec:stabilizing_cauchy}).
Furthermore, we expect based on
\citet{barton2017infinitesimal} and arguments above
that the distribution of segreation noise
is determined by the tail exponent of the distribution of effect sizes
(under appropriate assumptions; e.g., that most mutations are of small effect).
However, it is not clear how selection would affect these arguments,
especially in the stable case where some number of mutations are expected to
\emph{not} have small effect (and therefore such mutations will not be nearly neutral).

Therefore,
to study how this works in practice, we turn to simulation.
We used SLiM v4 \citep{haller2022slim4}
to simulate a discrete-time approximation to the continuous-time Moran model described above
as follows.
Individuals are diploid, hermaphroditic, and sexual;
the genome is of length $10^8$bp with a uniform recombination rate of $10^{-8}$ crossovers per bp;
mutations occur at a rate of $m$ per bp.
Each new mutation is independently assigned an ``effect''
drawn from the ``effect size distribution'' (and will be either Gaussian or Cauchy).
The phenotype of each individual is determined additively
by summing the effects of all mutations they carry across both chromosomes.
Let $f(z) = 1 + \mu(z)$
be the ``fitness'' of an individual with phenotype $z$,
and let $dt$ be a small constant (we took $dt=0.01$).
Then, at each time step
each individual dies with probability $1 - \exp(-dt f(z))$;
if there are $k$ such individuals chosen to die in a given time step,
then we choose $k$ individuals uniformly and without replacement
to produce new offspring;
for each such reproduction event the parent chooses a mate uniformly at random from the population.
This maintains the population at a fixed size, $N$.
(Note that individuals may self and individuals chosen to die may also reproduce,
but in a large population these details should be unimportant.)
